# Reported genotype landscape and database representation

```{r}
source("R/figures.R")
cohort_spectrum <- read_csv_or_example(
  "../results/thesis/cohort_reported_spectrum.csv",
  "data/example/cohort_reported_spectrum.csv"
)
```

## Reported diagnostic spectrum

The primary descriptive unit must be declared: person, test episode, reported
finding, allele, chromosome haplotype or diploid genotype. The positive fraction
in a referral cohort is not population prevalence.

```{r}
#| label: fig-reported-spectrum
#| fig-cap: "Example structure for the reported diagnostic spectrum."
#| fig-alt: "Horizontal bar chart of reported genotype categories."
plot_reported_spectrum(cohort_spectrum)
```

## Database representation

For each normalized allele or structural haplotype, compare nomenclature,
reference sequence, transcript, exact sequence/junction representation, clinical
assertions and population-resource coverage across IthaGenes/IthaCNVs, HbVar,
ClinVar and appropriate population resources.

## Haplotype and evolutionary context

A pathogenic HBB allele alone does not identify a classical beta-globin haplotype.
Individual-level founder or selection analyses require linked markers and phase.
Where those are absent, use external phased resources to describe known backgrounds
without assigning them to report-only patients.
