# Assay compiler, HBA family coordinates and genotype model

```{r}
source("R/figures.R")
assay_coverage <- read_tsv_or_example(
  "../results/assay/coverage.tsv",
  "data/example/assay_coverage.tsv"
)
molecule_status <- read_csv_or_example(
  "../results/thesis/molecule_status.csv",
  "data/example/molecule_status.csv"
)
```

## Forward assay model

```{mermaid}
flowchart LR
  G[Diploid chromosome-haplotype pair] --> P[Primer binding and compiled products]
  P --> A[Amplification efficiency and dropout]
  A --> O[Observed ONT molecules]
  O --> E[Primer, sequence, product and artifact evidence]
  E --> R[Ranked genotype classes]
```

## HBA family-coordinate and marker map

The family map is generated automatically by aligning HBA1, HBA2, hybrid and
candidate-copy sequences to one declared anchor. It records homologous reference
positions, insertion coordinates, gaps and copy-informative alleles. Adding HPRC
or cohort-derived sequences regenerates the map; it is not a manually curated
“full population equivalence map.”

Its purpose is to project copy markers and variants across representations and to
show when a read is copy-informative, family-informative or intrinsically
ambiguous. Exact product-sequence likelihoods remain the primary inference path.

## Assay coverage

```{r}
#| label: fig-assay-coverage
#| fig-cap: "Compiled product observability across candidate chromosome haplotypes."
#| fig-alt: "Tile plot of assay products by candidate haplotype, showing observable and absent products."
plot_assay_coverage(assay_coverage)
```

## Molecule admission

```{r}
#| label: fig-molecule-status
#| fig-cap: "Primer-aware classification of ONT reads."
#| fig-alt: "Horizontal bar chart of complete, incomplete, off-target and artifact read fractions."
plot_molecule_status(molecule_status)
```

## Candidate posterior

The model scores pairs of chromosome haplotypes. Variable alpha-globin copy number
is represented inside each chromosome-haplotype sequence; the human genotype
remains diploid. Named pairs with the same observable product signature are
collapsed before assigning the default prior, preventing catalogue duplication
from manufacturing posterior mass.
