# Evaluation

```{r}
source("R/figures.R")
posterior <- read_tsv_or_example(
  "../results/assay/genotypes/example.posteriors.tsv",
  "data/example/genotype_posteriors.tsv"
)
calibration_fit <- read_tsv_or_example(
  "../results/assay/calibration/sample_product_fit.tsv",
  "data/example/calibration_sample_product_fit.tsv"
)
backend_timings <- read_csv_or_example(
  "../results/benchmarks/genotype_backend_timings.csv",
  "data/example/genotype_backend_timings.csv"
)
parameter_stability <- read_csv_or_example(
  "../results/assay/stability/example.scenarios.csv",
  "data/example/genotype_parameter_stability.csv"
)
```

## Evaluation layers

Keep the following evaluations separate:

1. compiler correctness against known primer sites and expected products;
2. sealed synthetic genotype recovery and deliberate ambiguity fixtures;
3. real unlabelled amplicon mechanics, including endpoint, length and artifact
   distributions;
4. WGS small-variant benchmarking within declared truth/confident regions;
5. structural evidence compared with DRAGEN or graph calls as comparators;
6. control-based estimation of assay-product efficiency and dropout;
7. parameter-stability analysis around the fitted or declared model; and
8. analytical sensitivity/specificity on an independent certified or
   orthogonally characterised assay-matched set.

Calibration controls and final evaluation samples must be disjoint. The assay
profile, candidate catalogue and model parameters are frozen before blinded
analysis of the evaluation set.

## Posterior example

```{r}
#| label: fig-genotype-posterior
#| fig-cap: "Ranked conditional posterior over assay-observable genotype classes."
#| fig-alt: "Horizontal bar chart of the leading candidate genotype classes and their posterior probabilities."
plot_genotype_posterior(posterior)
```

## Calibration residuals

```{r}
#| label: fig-calibration-fit
#| fig-cap: "Observed and fitted effective product evidence in calibration controls."
#| fig-alt: "Faceted scatter plot comparing observed and fitted product weights with an identity line."
plot_calibration_fit(calibration_fit)
```

Residuals should be examined by product, genotype, run, lot and operator. A
well-fitting average can still conceal complete dropout for a clinically important
allele or one batch.

## Parameter stability

A posterior can be numerically sharp while depending heavily on an uncertain
product-dropout or artifact parameter. The declared grid therefore varies selected
scalar parameters, reruns the complete posterior and records both the top
observable class and the final call state.

```{r}
#| label: fig-parameter-stability
#| fig-cap: "Top-class posterior across declared dropout and unsupported-molecule parameter values."
#| fig-alt: "Heat map of top posterior over dropout and artifact probabilities, with percentages printed in cells."
plot_parameter_stability(parameter_stability)
```

Stability has two distinct meanings: the same biological class may remain top
ranked while its posterior drops below the research-call threshold, or the leading
class itself may change. Both the modal-class fraction and call-state distribution
should be reported. The grid is a sensitivity analysis, not a substitute for
control-based estimation.

## Execution performance

The full-sequence edit-distance path already uses a native RapidFuzz kernel. The
Cython implementation therefore targets grouped class-by-evidence scoring. Every
performance result first requires numerical agreement with the Python reference
for all likelihood terms and posterior probabilities.

```{r}
#| label: fig-backend-runtime
#| fig-cap: "Repeated wall-clock measurements for the Python reference and grouped Cython genotype scorers."
#| fig-alt: "Boxplots and individual timing points for Python and Cython backends."
plot_backend_timings(backend_timings)
```

Report the frozen input size, candidate haplotype count, observable class count,
read count, grouped evidence count, hardware, software commit, warm-up procedure,
and all repeated timings. Do not report only the fastest run.

## Ground-truth acquisition

The executable acquisition registry is maintained in `validation/resources.tsv`.
The immediate programme is:

- public multiplex-PCR reads for assay mechanics, not accuracy;
- HPRC and GIAB for sequence/caller questions after locus-confidence audit;
- the published Huang 2023 primer design as an executable public assay contract;
- requests for raw reads and truth tables from the Huang, Wei and Yang cohorts;
- orderable Coriell/CDC alpha- and beta-globin controls; and
- a blinded exact- AmplideX set paired with pre-existing orthogonal laboratory
  results.

DRAGEN and vendor Reporter calls remain comparators. They may enrich discordance
review but do not enter independent-truth accuracy denominators without separate
adjudication.

## Reporting

For each accuracy number, report the exact sample set, truth provenance, confidence
regions, variant class, TP/FP/FN denominators and comparison configuration. For the
amplicon model also report ambiguity, no-call and assay-not-observable rates. Do not
refer to DRAGEN-concordant deletion carriers as independent truth.
