# Evaluation

```{r}
source("R/figures.R")
posterior <- read_tsv_or_example(
  "../results/assay/genotypes/example.posteriors.tsv",
  "data/example/genotype_posteriors.tsv"
)
```

## Evaluation layers

Keep the following evaluations separate:

1. compiler correctness against known primer sites and expected products;
2. sealed synthetic genotype recovery and deliberate ambiguity fixtures;
3. real unlabelled amplicon mechanics, including endpoint, length and artifact
   distributions;
4. WGS small-variant benchmarking within declared truth/confident regions;
5. structural evidence compared with DRAGEN or graph calls as comparators; and
6. analytical sensitivity/specificity on certified or orthogonally characterised
   assay-matched material.

## Posterior example

```{r}
#| label: fig-genotype-posterior
#| fig-cap: "Ranked conditional posterior over assay-observable genotype classes."
#| fig-alt: "Horizontal bar chart of the leading candidate genotype classes and their posterior probabilities."
plot_genotype_posterior(posterior)
```

## Reporting

For each accuracy number, report the exact sample set, truth provenance, confidence
regions, variant class, TP/FP/FN denominators and comparison configuration. Do not
refer to DRAGEN-concordant deletion carriers as independent truth.
