# Evaluation

```{r}
source("R/figures.R")
small_variant_results <- read_required_tsv("../evaluation/small_variant_results.tsv")
hba_comparators <- read_required_tsv("../evaluation/hba_structural_comparators.tsv")
public_runs <- read_required_tsv("../evaluation/public_amplicon_runs.tsv")
cython_proof <- read_required_json("../benchmarks/cython_genotype_proof.json")
```

## Small-variant results

The committed result table contains six scoped evaluations. Five rows are the
original hap.py pool of HG002 plus five HPRC diploid-assembly genomes; the separate
HG002 row records a regional
GIAB `bcftools norm`/`isec` check. The original VCFs and per-sample hap.py output
are external to this checkout, so the committed records preserve counts, arithmetic,
truth source and comparison scope rather than pretending to recreate the benchmark
without its inputs.

```{r}
#| label: tbl-small-variant-results
#| tbl-cap: "Truth-scoped small-variant evaluation records."
small_variant_results |>
  transmute(
    Region = region,
    Type = variant_type,
    TP = tp,
    FN = fn,
    FP = fp,
    `Recall (%)` = round(100 * recall, 1),
    `Precision (%)` = round(100 * precision, 1)
  ) |>
  knitr::kable()
```

```{r}
#| label: fig-small-variant-performance
#| fig-cap: "Recall and precision calculated from the committed HPRC/GIAB counts."
#| fig-alt: "Point plot of recall and precision across six scoped globin evaluations."
plot_small_variant_performance(small_variant_results)
```

The strongest recorded result is beta-globin gene-body SNV performance: 156 TP,
5 FN and 1 FP. Indels are the main observed limitation. The 100% HG002 result
contains 23 regional small variants and does not establish whole-locus, structural,
or amplicon accuracy.

## HBA structural comparator checks

All committed structural rows are explicitly `comparator_only`.

```{r}
#| label: tbl-hba-structural-comparators
#| tbl-cap: "CIGAR-deletion observations against DRAGEN-labelled comparator states."
hba_comparators |>
  select(
    sample_id,
    comparator_label,
    observed_cigar_event_bp,
    deletion_read_fraction,
    interpretation,
    truth_role
  ) |>
  knitr::kable()
```

Agreement with DRAGEN is useful for debugging and discordance review, not independent
truth. Deletion size classes do not by themselves sequence-resolve every subtype;
large events have too few spanning molecules for reliable fraction-based zygosity;
and absence of a deletion signal in a gain comparator does not identify the exact
gain structure.

## Cython replacement gate

```{r}
#| label: tbl-cython-proof
#| tbl-cap: "Measured one-time replacement gate for the Cython genotype kernel."
tibble::tibble(
  `Candidate haplotypes` = cython_proof$workload$candidate_haplotypes,
  `Observable classes` = cython_proof$workload$observable_genotype_classes,
  `Molecule rows` = cython_proof$workload$molecule_rows,
  `Scalar median (s)` = cython_proof$historical_python_seconds$median,
  `Cython median (s)` = cython_proof$cython_seconds$median,
  `Median speedup` = cython_proof$median_speedup
) |>
  knitr::kable(digits = 3)
```

Numerical parity was established before the scalar implementation was removed. The
137.66-fold median improvement is a kernel result on a frozen workload, not an
end-to-end FASTQ performance claim.

## Public amplicon data and missing truth

The committed public-run registry contains `r dplyr::n_distinct(public_runs$run_accession)`
GridION accessions. These reads can support endpoint, length, product-imbalance,
artifact and runtime analyses. No independent sample-level genotype table is
committed for them, so they do not support genotype sensitivity or specificity.

A final analytical evaluation requires assay-matched raw reads paired with certified
reference genotypes or event-appropriate orthogonal evidence. Calibration controls
and the blinded evaluation set must remain disjoint.
